---
title: "Objective Sensitivity"
author: "inthomas@ethz.ch"
date: "`r Sys.Date()`"
format: 
  html:
    theme: cosmo
    toc: true
    code-fold: true
    code-tools: false
    fig-width: 12
    fig-height: 8
    lightbox: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(terra)
library(ggplot2)
library(dplyr)
library(patchwork)
library(reticulate)

# Configure Python environment
use_condaenv("rest_pymoo", required = TRUE)

# Set up Python imports
py_run_string("
import numpy as np
import rasterio as rio
from data_loader import load_initial_conditions
from resto_anom import recalculate_landscape_anomaly_with_conversions
import tempfile
import os
")

theme_set(theme_minimal())
```

## Load and Sample Initial Conditions

```{r load-data}
# Load full initial conditions
py_run_string("
initial_conditions = load_initial_conditions('.', objectives=['landscape'])
print('Loaded initial conditions with shape:', initial_conditions['shape'])
print('Available objectives:', [k for k in initial_conditions.keys() if 'anomaly' in k])
")

# Sample a 100x100 area from the center
py_run_string("
# Get original dimensions
full_shape = initial_conditions['shape']
center_r = full_shape[0] // 2
center_c = full_shape[1] // 2

# Define 100x100 sample area
sample_size = 100
start_r = max(0, center_r - sample_size // 2)
end_r = min(full_shape[0], start_r + sample_size)
start_c = max(0, center_c - sample_size // 2) 
end_c = min(full_shape[1], start_c + sample_size)

print(f'Sampling area: rows {start_r}:{end_r}, cols {start_c}:{end_c}')

# Sample the data
sample_shape = (end_r - start_r, end_c - start_c)
sampled_conditions = {}

# Sample spatial data
for key in ['landscape_anomaly', 'landscape_lulc_data', 'conversion_eligible_mask']:
    if key in initial_conditions:
        sampled_conditions[key] = initial_conditions[key][start_r:end_r, start_c:end_c]

# Sample other required data
sampled_conditions['shape'] = sample_shape
sampled_conditions['landscape_focal_classes'] = initial_conditions['landscape_focal_classes']
sampled_conditions['landscape_lulc_meta'] = initial_conditions['landscape_lulc_meta'].copy()

# Update metadata for sampled area
sampled_conditions['landscape_lulc_meta']['width'] = sample_shape[1]
sampled_conditions['landscape_lulc_meta']['height'] = sample_shape[0]
print(f'Sample shape: {sample_shape}')
")

# Extract data for R visualization
sample_shape <- py$sampled_conditions$shape
landscape_anomaly <- matrix(py$sampled_conditions$landscape_anomaly, nrow = sample_shape[[1]], ncol = sample_shape[[2]])
lulc_data <- matrix(py$sampled_conditions$landscape_lulc_data, nrow = sample_shape[[1]], ncol = sample_shape[[2]])
conversion_eligible <- matrix(as.numeric(py$sampled_conditions$conversion_eligible_mask), nrow = sample_shape[[1]], ncol = sample_shape[[2]])
focal_classes <- c(42, 43, 44, 45, 46, 47, 48, 49, 50, 51, 52, 53, 54, 55, 56, 57, 58, 59, 60, 64, 65, 66, 67)

cat("Sample area dimensions:", dim(landscape_anomaly), "\n")
cat("Focal classes for conversion:", focal_classes, "\n")
cat("Eligible pixels for conversion:", sum(conversion_eligible), "\n")
```

## Visualize Initial Conditions

```{r visualize-initial}
# Convert to data frames for ggplot
create_plot_df <- function(data, name) {
  df <- expand.grid(x = 1:ncol(data), y = 1:nrow(data))
  df$value <- as.vector(data)
  df$type <- name
  df$y <- max(df$y) - df$y + 1  # Flip y-axis for proper orientation
  return(df)
}

# Create plotting data
anomaly_df <- create_plot_df(landscape_anomaly, "Landscape Anomaly")
lulc_df <- create_plot_df(lulc_data, "LULC Classes")
eligible_df <- create_plot_df(conversion_eligible, "Conversion Eligible")

# Plot initial landscape anomaly
p1 <- ggplot(anomaly_df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Anomaly") +
  labs(title = "Initial Landscape Anomaly", 
       subtitle = paste("Mean anomaly:", round(mean(landscape_anomaly, na.rm=TRUE), 3))) +
  coord_equal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
p1
# Plot LULC classes
p2 <- ggplot(lulc_df, aes(x = x, y = y, fill = factor(value))) +
  geom_raster() +
  scale_fill_viridis_d(name = "LULC Class") +
  labs(title = "Land Use Land Cover",
       subtitle = paste("Unique classes:", length(unique(as.vector(lulc_data))))) +
  coord_equal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
p2
# Plot conversion eligible areas
p3 <- ggplot(eligible_df, aes(x = x, y = y, fill = factor(value))) +
  geom_raster() +
  scale_fill_manual(values = c("0" = "lightgray", "1" = "red"), 
                   name = "Eligible", labels = c("No", "Yes")) +
  labs(title = "Conversion Eligible Pixels",
       subtitle = paste("Eligible pixels:", sum(conversion_eligible))) +
  coord_equal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
p3

```

## Generate Random Conversion Patterns

```{r random-conversions}
# Test different conversion intensities
conversion_fractions <- c(0.05, 0.10, 0.15, 0.20)
results_list <- list()

for (i in seq_along(conversion_fractions)) {
  frac <- conversion_fractions[i]
  
  # Generate random conversion pattern in Python
  py_run_string(sprintf("
# Generate random conversion decisions
conversion_frac = %f
eligible_pixels = np.sum(sampled_conditions['conversion_eligible_mask'])
n_convert = int(conversion_frac * eligible_pixels)

print(f'Converting {n_convert} pixels ({conversion_frac*100:.1f}%% of {eligible_pixels} eligible)')

# Create random conversion mask
conversion_mask_2d = np.zeros(sampled_conditions['shape'], dtype=bool)
if n_convert > 0:
    # Get eligible pixel coordinates
    eligible_coords = np.where(sampled_conditions['conversion_eligible_mask'])
    eligible_indices = list(zip(eligible_coords[0], eligible_coords[1]))
    
    # Randomly select pixels to convert
    np.random.seed(42 + %d)  # Different seed for each fraction
    selected_indices = np.random.choice(len(eligible_indices), size=n_convert, replace=False)
    
    for idx in selected_indices:
        r, c = eligible_indices[idx]
        conversion_mask_2d[r, c] = True

print(f'Conversion mask sum: {np.sum(conversion_mask_2d)}')

# Calculate updated landscape anomaly
try:
    updated_anomaly = recalculate_landscape_anomaly_with_conversions(
        sampled_conditions, conversion_mask_2d
    )
    
    # Calculate landscape objective value using the same formula as resto_anom.py
    eps = 1e-12
    landscape_objective = np.sum(updated_anomaly - sampled_conditions['landscape_anomaly']) / (np.sum(sampled_conditions['landscape_anomaly']) + eps)
    
    print(f'Initial landscape anomaly sum: {np.sum(sampled_conditions[\"landscape_anomaly\"]):.6f}')
    print(f'Updated landscape anomaly sum: {np.sum(updated_anomaly):.6f}')
    print(f'Landscape objective value: {landscape_objective:.6f}')
    
    conversion_result = {
        'fraction': conversion_frac,
        'n_converted': n_convert,
        'landscape_objective': landscape_objective,
        'conversion_mask': conversion_mask_2d,
        'updated_anomaly': updated_anomaly
    }
except Exception as e:
    print(f'Error in calculation: {e}')
    conversion_result = None
  ", frac, i))
  
  # Extract results
  if (!is.null(py$conversion_result)) {
    result <- list(
      fraction = py$conversion_result$fraction,
      n_converted = py$conversion_result$n_converted,
      landscape_objective = py$conversion_result$landscape_objective,
      conversion_mask = py$conversion_result$conversion_mask,
      updated_anomaly = py$conversion_result$updated_anomaly
    )
    results_list[[i]] <- result
  }
}

# Display results summary
results_df <- do.call(rbind, lapply(results_list, function(r) {
  data.frame(
    Conversion_Fraction = r$fraction,
    Pixels_Converted = r$n_converted,
    Landscape_Objective = r$landscape_objective
  )
}))

knitr::kable(results_df, digits = 6, 
             caption = "Landscape Objective Response to Conversion Intensity")
```

```{r fix-conversion-logic}
# Check what's actually happening with focal classes
py_run_string("
print('=== FOCAL CLASS ANALYSIS ===')
focal_classes_full = [42, 43, 44, 45, 46, 47, 48, 49, 50, 51,
                     52, 53, 54, 55, 56, 57, 58, 59, 60, 64, 65, 66, 67]
exclude_classes = np.concatenate([
    np.arange(1, 22),          # 1–21  
    np.array([61, 62, 63])
])
focal_classes_final = np.setdiff1d(focal_classes_full, exclude_classes)

print(f'Original focal classes: {focal_classes_full}')
print(f'Exclude classes: {exclude_classes}') 
print(f'Final focal classes (after exclusion): {focal_classes_final}')

# Check LULC values in sample
lulc_unique = np.unique(sampled_conditions['landscape_lulc_data'])
print(f'Unique LULC values in sample: {lulc_unique}')

# Check which are eligible for conversion
eligible_values = lulc_unique[~np.isin(lulc_unique, focal_classes_final) & 
                             ~np.isin(lulc_unique, exclude_classes)]
print(f'LULC values eligible for conversion: {eligible_values}')

# Store for demo
sampled_conditions['focal_classes_final'] = focal_classes_final
sampled_conditions['exclude_classes'] = exclude_classes
sampled_conditions['eligible_lulc_values'] = eligible_values
")

# Update R variables
focal_classes_final <- py$sampled_conditions$focal_classes_final
exclude_classes <- py$sampled_conditions$exclude_classes
eligible_lulc_values <- py$sampled_conditions$eligible_lulc_values

cat("Focal classes (final):", focal_classes_final, "\n")
cat("Exclude classes:", exclude_classes, "\n") 
cat("Eligible LULC values for conversion:", eligible_lulc_values, "\n")
```



```{r corrected-conversion-test}
# Test conversion with proper before/after LULC tracking
py_run_string("
# Test single conversion to verify logic
conversion_frac = 0.10
eligible_pixels = np.sum(sampled_conditions['conversion_eligible_mask'])
n_convert = int(conversion_frac * eligible_pixels)

print(f'=== CONVERSION TEST ===')
print(f'Converting {n_convert} pixels ({conversion_frac*100:.1f}% of {eligible_pixels} eligible)')

# Create conversion mask
conversion_mask_2d = np.zeros(sampled_conditions['shape'], dtype=bool)
if n_convert > 0:
    eligible_coords = np.where(sampled_conditions['conversion_eligible_mask'])
    eligible_indices = list(zip(eligible_coords[0], eligible_coords[1]))
    
    np.random.seed(42)
    selected_indices = np.random.choice(len(eligible_indices), size=min(n_convert, len(eligible_indices)), replace=False)
    
    # Track before/after LULC values
    before_lulc_values = []
    for idx in selected_indices:
        r, c = eligible_indices[idx]
        conversion_mask_2d[r, c] = True
        before_lulc_values.append(sampled_conditions['landscape_lulc_data'][r, c])
    
    print(f'LULC values before conversion: {np.unique(before_lulc_values)}')

# Apply conversion and check results
try:
    updated_anomaly = recalculate_landscape_anomaly_with_conversions(
        sampled_conditions, conversion_mask_2d
    )
    
    # The conversion function should have modified the LULC - but we need to see what it converted TO
    print('Conversion successful - landscape anomaly recalculated')
    
    conversion_result_test = {
        'conversion_mask': conversion_mask_2d,
        'updated_anomaly': updated_anomaly,
        'before_lulc_values': before_lulc_values
    }
except Exception as e:
    print(f'Conversion failed: {e}')
    conversion_result_test = None
")
```

## Visualize Changes
```{r visualize-changes}
# Plot conversion patterns and resulting changes
plot_list <- list()

for (i in seq_along(results_list)) {
  result <- results_list[[i]]
  
  # Convert boolean mask to numeric matrix for R plotting
  conversion_mask_numeric <- matrix(as.numeric(result$conversion_mask), 
                                   nrow = sample_shape[[1]], 
                                   ncol = sample_shape[[2]])
  
  # Convert updated anomaly to matrix
  updated_anomaly_matrix <- matrix(result$updated_anomaly,
                                  nrow = sample_shape[[1]], 
                                  ncol = sample_shape[[2]])
  
  # Create difference map
  difference <- updated_anomaly_matrix - landscape_anomaly

  # Create plot data
  conv_df <- create_plot_df(conversion_mask_numeric, "Conversion")
  diff_df <- create_plot_df(difference, "Difference")
  
  # Conversion pattern plot
  p_conv <- ggplot(conv_df, aes(x = x, y = y, fill = factor(value))) +
    geom_raster() +
    scale_fill_manual(values = c("0" = "lightgray", "1" = "red"),
                     labels = c("0" = "Not converted", "1" = "Converted"),
                     name = "") +
    labs(title = paste0("Conversion Pattern (", result$fraction*100, "%)"),
         subtitle = paste(result$n_converted, "pixels converted")) +
    coord_equal() +
    theme_void() +
    theme(legend.position = "bottom")
  
  # Difference plot
  p_diff <- ggplot(diff_df, aes(x = x, y = y, fill = value)) +
    geom_raster() +
    scale_fill_gradient2(name = "Δ Anomaly", 
                        low = "blue", mid = "white", high = "red",
                        midpoint = 0) +
    labs(title = "Change in Landscape Anomaly",
         subtitle = paste("Landscape objective:", sprintf("%.6f", result$landscape_objective))) +
    coord_equal() +
    theme_void() +
    theme(legend.position = "bottom")
  
  plot_list[[i]] <- p_conv | p_diff
}

# Display all conversion scenarios
for (i in seq_along(plot_list)) {
  print(plot_list[[i]])
}
```

## Sensitivity to number of pixels converted

```{r summary-plot}
# Create sensitivity curve
sensitivity_df <- data.frame(
  Conversion_Fraction = sapply(results_list, function(r) r$fraction * 100),
  Landscape_Objective = sapply(results_list, function(r) r$landscape_objective),
  Pixels_Converted = sapply(results_list, function(r) r$n_converted)
)

# Landscape objective response plot
p_obj <- ggplot(sensitivity_df, aes(x = Conversion_Fraction, y = Landscape_Objective)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "darkblue") +
  labs(title = "Landscape Objective Response",
       subtitle = "Objective value as calculated in resto_anom.py",
       x = "Conversion Fraction (%)",
       y = "Landscape Objective Value") +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5)

# Objective vs pixel count
p_pixels <- ggplot(sensitivity_df, aes(x = Pixels_Converted, y = Landscape_Objective)) +
  geom_line(size = 1.2, color = "darkgreen") +
  geom_point(size = 3, color = "forestgreen") +
  labs(title = "Objective vs. Conversion Intensity",
       x = "Number of Pixels Converted",
       y = "Landscape Objective Value") +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5)

# Combine sensitivity plots
sensitivity_plots <- p_obj / p_pixels
print(sensitivity_plots)

# Display final summary
cat("\n## Key Findings\n\n")
cat("- **Baseline landscape anomaly sum:**", sprintf("%.6f", sum(landscape_anomaly)), "\n")
cat("- **Maximum tested conversion:**", max(sensitivity_df$Conversion_Fraction), "% of eligible pixels\n")
cat("- **Landscape objective range:**", sprintf("%.6f to %.6f", min(sensitivity_df$Landscape_Objective), max(sensitivity_df$Landscape_Objective)), "\n")
cat("- **Best objective value:**", sprintf("%.6f", min(sensitivity_df$Landscape_Objective)), "(lowest value is best)\n")

# Calculate sensitivity per pixel
sensitivity_per_pixel <- sensitivity_df$Landscape_Objective / sensitivity_df$Pixels_Converted
cat("- **Average objective change per pixel:**", sprintf("%.6f", mean(sensitivity_per_pixel, na.rm=TRUE)), "\n")
```

## Optimization Signal Analysis

This analysis tests whether restoration and conversion decisions produce meaningful objective variation - essential for effective optimization.

```{r optimization-signal-check}
# Test optimization signal using sample solutions
py_run_string("
print('=== OPTIMIZATION SIGNAL CHECK ===')

# Create a simple problem setup for signal testing
scenario_params_test = {
    'max_restoration_fraction': 0.2,
    'spatial_clustering': 0.0,
    'burden_sharing': 'no',
    'abiotic_effect': 0.01,
    'biotic_effect': 0.01,
    'anomaly_weight_shape': 'exponential',
    'anomaly_weight_scale': 1.0
}

# Load initial conditions for all objectives
try:
    initial_conditions_test = load_initial_conditions('.', objectives=['abiotic', 'biotic', 'landscape', 'cost'])
    print(f'Loaded objectives: {[k for k in initial_conditions_test.keys() if \"anomaly\" in k or k == \"implementation_cost\"]}')
    print(f'Restoration eligible pixels: {initial_conditions_test[\"n_restoration_pixels\"]}')
    print(f'Conversion eligible pixels: {initial_conditions_test[\"n_conversion_pixels\"]}')
    
    # Quick signal check: test different action fractions
    test_fractions = [0.05, 0.15, 0.25, 0.35]
    n_samples = 5
    conv_share = 0.5  # Share actions between restoration (0.5) and conversion (0.5)
    
    # Determine available objectives
    objective_names = []
    if 'abiotic_anomaly' in initial_conditions_test:
        objective_names.append('abiotic_anomaly')
    if 'biotic_anomaly' in initial_conditions_test:
        objective_names.append('biotic_anomaly')
    if 'landscape_anomaly' in initial_conditions_test:
        objective_names.append('landscape_anomaly')
    if 'implementation_cost' in initial_conditions_test:
        objective_names.append('implementation_cost')
    
    print(f'Testing objectives: {objective_names}')
    
    # Calculate action budgets based on eligible pixels
    n_restoration_pixels = initial_conditions_test['n_restoration_pixels']
    n_conversion_pixels = initial_conditions_test['n_conversion_pixels']
    max_restored_pixels = int(scenario_params_test['max_restoration_fraction'] * n_restoration_pixels)
    
    signal_results = {}
    
    for obj_name in objective_names:
        print(f'\\nTesting {obj_name}...')
        objective_values = []
        
        for frac in test_fractions:
            # Calculate number of actions for this fraction
            n_actions = int(frac * max_restored_pixels)
            
            for sample in range(n_samples):
                np.random.seed(42 + sample + int(frac*100))  # Reproducible seeds
                
                # Split actions between restoration and conversion
                n_conv = int(round(conv_share * n_actions))
                n_rest = n_actions - n_conv
                
                # Ensure we don't exceed available pixels
                n_rest = min(n_rest, n_restoration_pixels)
                n_conv = min(n_conv, n_conversion_pixels)
                
                # Create decision vectors
                restore_vars = np.zeros(n_restoration_pixels, dtype=int)
                convert_vars = np.zeros(n_conversion_pixels, dtype=int)
                
                if n_rest > 0:
                    rest_idx = np.random.permutation(n_restoration_pixels)[:n_rest]
                    restore_vars[rest_idx] = 1
                
                if n_conv > 0:
                    conv_idx = np.random.permutation(n_conversion_pixels)[:n_conv]
                    convert_vars[conv_idx] = 1
                
                # Calculate objective value using restoration_effect function
                updated_conditions = restoration_effect(restore_vars, convert_vars, initial_conditions_test, scenario_params_test)
                
                # Calculate objective based on type
                if obj_name in ['abiotic_anomaly', 'biotic_anomaly']:
                    # Maximize improvement relative to baseline (minimize negative sum)
                    base = initial_conditions_test[obj_name]
                    mask = initial_conditions_test['restoration_eligible_mask']
                    obj_value = -np.sum((updated_conditions[obj_name] - base)[mask])
                    
                elif obj_name == 'landscape_anomaly':
                    # Relative change in landscape anomaly
                    l0 = initial_conditions_test['landscape_anomaly']
                    l1 = updated_conditions['landscape_anomaly']
                    eps = 1e-12
                    obj_value = np.sum(l1 - l0) / (np.sum(l0) + eps)
                    
                elif obj_name == 'implementation_cost':
                    # Total implementation cost
                    obj_value = updated_conditions[obj_name]
                else:
                    obj_value = 0.0
                
                objective_values.append(obj_value)
                
                if sample == 0 and frac == test_fractions[0]:  # Debug first sample
                    print(f'  Sample: {n_rest} restore + {n_conv} convert = {n_rest + n_conv} total actions')
                    print(f'  Objective value: {obj_value:.6f}')
        
        # Calculate signal statistics
        if len(objective_values) > 1:
            obj_range = float(np.max(objective_values) - np.min(objective_values))
            mean_val = float(np.mean(objective_values))
            std_val = float(np.std(objective_values))
            cv = std_val / abs(mean_val) if mean_val != 0 else 0.0
            
            signal_results[obj_name] = {
                'range': obj_range,
                'mean': mean_val,
                'std': std_val,
                'cv': cv,
                'values': objective_values
            }
            
            print(f'  Range: {obj_range:.2e}')
            print(f'  Mean: {mean_val:.6f}')
            print(f'  CV (std/mean): {cv:.3f}')
        else:
            print(f'  No values collected for {obj_name}')
            signal_results[obj_name] = {'range': 0, 'mean': 0, 'std': 0, 'cv': 0, 'values': []}
    
    # Overall signal assessment
    cv_threshold = 0.1
    has_signal = any(result['cv'] > cv_threshold for result in signal_results.values())
    
    print(f'\\n=== SIGNAL ASSESSMENT ===')
    print(f'CV threshold for meaningful signal: {cv_threshold}')
    print(f'Optimization signal: {\"✓ PRESENT\" if has_signal else \"✗ WEAK or ABSENT\"}')
    
    # Store results for R visualization
    signal_analysis = {
        'results': signal_results,
        'has_signal': has_signal,
        'cv_threshold': cv_threshold,
        'test_parameters': {
            'test_fractions': test_fractions,
            'n_samples': n_samples,
            'conv_share': conv_share,
            'max_restored_pixels': max_restored_pixels
        }
    }
    
except Exception as e:
    print(f'Error in signal analysis: {e}')
    signal_analysis = None
")

# Extract and display results
if (!is.null(py$signal_analysis)) {
  # Create signal summary table
  signal_summary <- do.call(rbind, lapply(names(py$signal_analysis$results), function(obj_name) {
    result <- py$signal_analysis$results[[obj_name]]
    data.frame(
      Objective = obj_name,
      Range = result$range,
      Mean = result$mean,
      Std_Dev = result$std,
      CV = result$cv,
      Signal = ifelse(result$cv > py$signal_analysis$cv_threshold, "Strong", "Weak")
    )
  }))
  
  knitr::kable(signal_summary, digits = 6, 
               caption = "Optimization Signal Analysis - Objective Variation Test")
  
  # Create visualization of objective variation
  if (nrow(signal_summary) > 0) {
    library(ggplot2)
    
    # CV plot
    p_signal <- ggplot(signal_summary, aes(x = Objective, y = CV, fill = Signal)) +
      geom_col(alpha = 0.7) +
      geom_hline(yintercept = py$signal_analysis$cv_threshold, linetype = "dashed", color = "red") +
      scale_fill_manual(values = c("Strong" = "darkgreen", "Weak" = "orange")) +
      labs(title = "Optimization Signal Strength",
           subtitle = paste("CV >", py$signal_analysis$cv_threshold, "indicates meaningful objective variation"),
           x = "Objective", y = "Coefficient of Variation (CV)") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
    print(p_signal)
    
    # Signal assessment
    if (py$signal_analysis$has_signal) {
      cat("\n✓ **OPTIMIZATION SIGNAL PRESENT**\n")
      cat("Objectives show meaningful variation with different restoration/conversion decisions.\n")
      cat("Optimization should be able to find diverse trade-off solutions.\n")
    } else {
      cat("\n⚠ **WEAK OPTIMIZATION SIGNAL**\n") 
      cat("Limited objective variation detected. Consider:\n")
      cat("- Adjusting restoration effect parameters\n")
      cat("- Increasing action budgets\n")
      cat("- Reviewing objective formulations\n")
    }
    
    # Test parameters summary
    cat("\n**Test Parameters:**\n")
    cat("- Action fractions tested:", paste(py$signal_analysis$test_parameters$test_fractions, collapse = ", "), "\n")
    cat("- Samples per fraction:", py$signal_analysis$test_parameters$n_samples, "\n")
    cat("- Conversion share:", py$signal_analysis$test_parameters$conv_share, "\n")
    cat("- Max restored pixels:", py$signal_analysis$test_parameters$max_restored_pixels, "\n")
  }
  
} else {
  cat("Signal analysis failed - check Python output for errors.\n")
}
```


## Marginal Gain Surfaces for Restoration

Based on the abiotic and biotic objective calculations in resto_anom.py, we can calculate the per-pixel marginal gain for restoration actions.

```{r marginal-gain-analysis}
# Calculate marginal gain surfaces for abiotic and biotic restoration
py_run_string("
print('=== MARGINAL GAIN CALCULATION ===')

# Load full initial conditions with abiotic and biotic data
initial_conditions_full = load_initial_conditions('.', objectives=['abiotic', 'biotic'])

# Sample the same area as landscape data
full_shape = initial_conditions_full['shape']
center_r = full_shape[0] // 2
center_c = full_shape[1] // 2

# Use same sampling area as landscape demo
sample_size = 100
start_r = max(0, center_r - sample_size // 2)
end_r = min(full_shape[0], start_r + sample_size)
start_c = max(0, center_c - sample_size // 2) 
end_c = min(full_shape[1], start_c + sample_size)

# Sample abiotic and biotic data
abiotic_anomaly_sample = initial_conditions_full['abiotic_anomaly'][start_r:end_r, start_c:end_c]
biotic_anomaly_sample = initial_conditions_full['biotic_anomaly'][start_r:end_r, start_c:end_c]
restoration_eligible_sample = initial_conditions_full['restoration_eligible_mask'][start_r:end_r, start_c:end_c]

print(f'Sampled abiotic anomaly shape: {abiotic_anomaly_sample.shape}')
print(f'Sampled biotic anomaly shape: {biotic_anomaly_sample.shape}')
print(f'Restoration eligible pixels in sample: {np.sum(restoration_eligible_sample)}')

# Implement the anomaly improvement weight function (from resto_anom.py)
def anomaly_improvement_weight(anomaly_values, shape='exponential', scale=1.0):
    neg_mask = anomaly_values < 0
    abs_anomaly = np.abs(anomaly_values)
    
    if shape == 'exponential':
        weights = 1 - np.exp(-abs_anomaly / scale)
    elif shape == 'gaussian':
        weights = np.exp(-(abs_anomaly**2) / (2 * scale**2))
    else:
        raise ValueError(f'Unknown weight shape: {shape}')
    
    gamma = 3.0
    weights = weights ** gamma
    weights = np.where(neg_mask, weights, 0.0)
    weights = np.clip(weights, 0.0, 1.0)
    
    return weights

# Default restoration effect parameters (from resto_anom.py defaults)
abiotic_effect = 0.01  # Default abiotic improvement
biotic_effect = 0.01   # Default biotic improvement
weight_shape = 'exponential'
weight_scale = 1.0

# Calculate improvement weights for each pixel
abiotic_weights = anomaly_improvement_weight(abiotic_anomaly_sample, shape=weight_shape, scale=weight_scale)
biotic_weights = anomaly_improvement_weight(biotic_anomaly_sample, shape=weight_shape, scale=weight_scale)

# Calculate marginal gain per pixel (improvement * weight)
abiotic_marginal_gain = abiotic_effect * abiotic_weights
biotic_marginal_gain = biotic_effect * biotic_weights

# Apply restoration eligibility mask (only restoration-eligible pixels can have gains)
abiotic_marginal_gain = np.where(restoration_eligible_sample, abiotic_marginal_gain, 0.0)
biotic_marginal_gain = np.where(restoration_eligible_sample, biotic_marginal_gain, 0.0)

# Store results for R visualization
marginal_gain_data = {
    'abiotic_anomaly': abiotic_anomaly_sample,
    'biotic_anomaly': biotic_anomaly_sample,
    'abiotic_weights': abiotic_weights,
    'biotic_weights': biotic_weights,
    'abiotic_marginal_gain': abiotic_marginal_gain,
    'biotic_marginal_gain': biotic_marginal_gain,
    'restoration_eligible': restoration_eligible_sample,
    'sample_shape': (end_r - start_r, end_c - start_c),
    'effects': {'abiotic': abiotic_effect, 'biotic': biotic_effect}
}

print(f'Abiotic marginal gain range: {np.min(abiotic_marginal_gain):.6f} to {np.max(abiotic_marginal_gain):.6f}')
print(f'Biotic marginal gain range: {np.min(biotic_marginal_gain):.6f} to {np.max(biotic_marginal_gain):.6f}')
print(f'Non-zero abiotic gains: {np.sum(abiotic_marginal_gain > 0)} pixels')
print(f'Non-zero biotic gains: {np.sum(biotic_marginal_gain > 0)} pixels')
")

# Extract data for R visualization
sample_shape_mg <- py$marginal_gain_data$sample_shape
abiotic_anomaly <- matrix(py$marginal_gain_data$abiotic_anomaly, nrow = sample_shape_mg[[1]], ncol = sample_shape_mg[[2]])
biotic_anomaly <- matrix(py$marginal_gain_data$biotic_anomaly, nrow = sample_shape_mg[[1]], ncol = sample_shape_mg[[2]])
abiotic_marginal_gain <- matrix(py$marginal_gain_data$abiotic_marginal_gain, nrow = sample_shape_mg[[1]], ncol = sample_shape_mg[[2]])
biotic_marginal_gain <- matrix(py$marginal_gain_data$biotic_marginal_gain, nrow = sample_shape_mg[[1]], ncol = sample_shape_mg[[2]])
restoration_eligible <- matrix(as.numeric(py$marginal_gain_data$restoration_eligible), nrow = sample_shape_mg[[1]], ncol = sample_shape_mg[[2]])

cat("Marginal gain analysis completed:\n")
cat("- Abiotic effect:", py$marginal_gain_data$effects$abiotic, "\n")
cat("- Biotic effect:", py$marginal_gain_data$effects$biotic, "\n")
cat("- Eligible pixels for restoration:", sum(restoration_eligible), "\n")
```

### Visualize Marginal Gain Surfaces

```{r visualize-marginal-gains}
# Create plotting data for marginal gains
abiotic_anomaly_df <- create_plot_df(abiotic_anomaly, "Abiotic Anomaly")
biotic_anomaly_df <- create_plot_df(biotic_anomaly, "Biotic Anomaly")
abiotic_gain_df <- create_plot_df(abiotic_marginal_gain, "Abiotic Marginal Gain")
biotic_gain_df <- create_plot_df(biotic_marginal_gain, "Biotic Marginal Gain")
restoration_eligible_df <- create_plot_df(restoration_eligible, "Restoration Eligible")

# Plot baseline anomalies
p_abiotic_anomaly <- ggplot(abiotic_anomaly_df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Anomaly", option = "plasma") +
  labs(title = "Abiotic Anomaly (Baseline)", 
       subtitle = paste("Range:", round(min(abiotic_anomaly), 3), "to", round(max(abiotic_anomaly), 3))) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "bottom")

p_biotic_anomaly <- ggplot(biotic_anomaly_df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Anomaly", option = "plasma") +
  labs(title = "Biotic Anomaly (Baseline)", 
       subtitle = paste("Range:", round(min(biotic_anomaly), 3), "to", round(max(biotic_anomaly), 3))) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "bottom")

# Plot marginal gain surfaces
p_abiotic_gain <- ggplot(abiotic_gain_df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Marginal\nGain", option = "viridis", 
                       trans = "sqrt",  # Square root transformation to show detail in low values
                       labels = function(x) sprintf("%.4f", x)) +
  labs(title = "Abiotic Marginal Gain per Pixel", 
       subtitle = paste("Max gain:", sprintf("%.4f", max(abiotic_marginal_gain)), 
                       "| Non-zero pixels:", sum(abiotic_marginal_gain > 0))) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "bottom")

p_biotic_gain <- ggplot(biotic_gain_df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Marginal\nGain", option = "viridis",
                       trans = "sqrt",  # Square root transformation to show detail in low values
                       labels = function(x) sprintf("%.4f", x)) +
  labs(title = "Biotic Marginal Gain per Pixel", 
       subtitle = paste("Max gain:", sprintf("%.4f", max(biotic_marginal_gain)), 
                       "| Non-zero pixels:", sum(biotic_marginal_gain > 0))) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "bottom")

# Plot restoration eligibility
p_restoration_eligible <- ggplot(restoration_eligible_df, aes(x = x, y = y, fill = factor(value))) +
  geom_raster() +
  scale_fill_manual(values = c("0" = "lightgray", "1" = "darkgreen"), 
                   name = "Eligible", labels = c("No", "Yes")) +
  labs(title = "Restoration Eligible Pixels",
       subtitle = paste("Eligible pixels:", sum(restoration_eligible), "of", length(restoration_eligible))) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "bottom")

# Combine plots
baseline_plots <- p_abiotic_anomaly | p_biotic_anomaly | p_restoration_eligible
gain_plots <- p_abiotic_gain | p_biotic_gain

# Display plots
print(baseline_plots)
print(gain_plots)
```

### Marginal Gain Statistics

```{r marginal-gain-stats}
# Calculate statistics for marginal gains (only for eligible pixels)
eligible_mask <- restoration_eligible == 1

if (sum(eligible_mask) > 0) {
  # Extract values for eligible pixels only
  abiotic_gains_eligible <- abiotic_marginal_gain[eligible_mask]
  biotic_gains_eligible <- biotic_marginal_gain[eligible_mask]
  
  # Create summary statistics
  gain_stats <- data.frame(
    Objective = c("Abiotic", "Biotic"),
    Effect_Parameter = c(py$marginal_gain_data$effects$abiotic, py$marginal_gain_data$effects$biotic),
    Mean_Gain = c(mean(abiotic_gains_eligible), mean(biotic_gains_eligible)),
    Median_Gain = c(median(abiotic_gains_eligible), median(biotic_gains_eligible)),
    Max_Gain = c(max(abiotic_gains_eligible), max(biotic_gains_eligible)),
    Min_Gain = c(min(abiotic_gains_eligible), min(biotic_gains_eligible)),
    SD_Gain = c(sd(abiotic_gains_eligible), sd(biotic_gains_eligible)),
    Eligible_Pixels = c(sum(abiotic_gains_eligible > 0), sum(biotic_gains_eligible > 0))
  )
  
  knitr::kable(gain_stats, digits = 6, 
               caption = "Marginal Gain Statistics for Restoration-Eligible Pixels")
  
  # Create histograms of marginal gains
  gain_data <- data.frame(
    Gain = c(abiotic_gains_eligible, biotic_gains_eligible),
    Objective = rep(c("Abiotic", "Biotic"), each = length(abiotic_gains_eligible))
  )
  
  p_gain_hist <- ggplot(gain_data, aes(x = Gain, fill = Objective)) +
    geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
    scale_fill_viridis_d(option = "plasma", end = 0.8) +
    labs(title = "Distribution of Marginal Gains",
         subtitle = "For restoration-eligible pixels only",
         x = "Marginal Gain per Pixel", y = "Frequency") +
    facet_wrap(~Objective, scales = "free") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  print(p_gain_hist)
  
} else {
  cat("No restoration-eligible pixels found in sample area.\n")
}
```

- Marginal gain represents the objective improvement achieved by restoring each individual pixel
- Calculated as: `restoration_effect × anomaly_improvement_weight(baseline_anomaly)`
- Only applies to restoration-eligible pixels (ecosystem-specific masks)
- Negative anomalies get higher weights: Pixels with more negative baseline anomalies receive larger restoration improvements